Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-16 edge/model.tex
reorganize the model material

\section{Settings and learning problem}
\label{sec:edge_model}

As in the previous chapters, we are given an unweighted graph $G=(V,E)$\stodo{undirected?}. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator between feature vectors $s$ such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$\stodo{is it better to
introduce $w$ with a single $\ell$ index, to emphasise that later, it will not be unique per edge?},
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm. As we will discuss, this
constraint can be seen as a way to couple the different dimensions of the profiles.

@@

We assume in the following that both the profiles and the directions are vectors with components in
$[-1, 1]$. A natural choice for the operator $s$ is then the difference between the profile, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_{uv}$, we expect
${x_u}^T w_{uv}$ to be close to ${x_v}^T w_{uv}$, as measured by $|{x_u}^T w_{uv} - {x_v}^T
w_{uv}|$. Because we prefer to work with smooth functions, we define the goodness to be $g(s_{uv},
w_{uv}) = \left({s_{uv}}^T w_{uv} \right)^2$, and we want to minimize it with respect to $w_{uv}$.
When ${x_u}_{;i} - {x_v}_{;i}$ is close to zero, the minimization and the fact that $w_{uv}$ is
unit-norm impose that ${w_{uv}}_{;i}$ should be large. On the other hand, with this formulation,
there is no way to set ${w_{uv}}_{;i}$ in order to \enquote{reward/favor} the fact that
${x_u}_{;i}$ and ${x_v}_{;i}$ are widely different.
Whereas there might be situations where this is not a problem, as we discussed in the introduction,
here we seek to model heterophilic interactions, at least along a subset of dimensions. Therefore, in
the following, we will adopt a refined approach.

@@

\begin{aside}
When $s$ is a distance, and therefore positive, we impose $w_{uv}\geq 0$. A link $(u,v)$ is then
well explained by $w_{uv}$ when observed differences along this direction are small. More precisely,
when the quantity ${s_{uv}}^T w_{uv}$ is close to zero. Because $w_{uv}$ is unit norm, the \ith{}
component of $w_{uv}$ should be large when the \ith{} feature of $x_u$ and $x_v$ are close. Hence, a
null value for a component $i$ of $s_{uv}$ should not mean anything else than $u$ and $v$ share a
known and similar feature $i$ (\eg{} the unknown value should be different from $0$). When $s$ is
the difference $x_u - x_v$, it can be negative and the same reasoning can be applied, but with
$w_{uv}\in [-1,1]^d$ and a link value of $\left({s_{uv}}^T w_{uv}\right)^2$.

An alternative is to consider that the combination is a similarity. Links are created when the
observed similarity is high. Again, if $s$ is positive, it seems reasonable to impose that
$w_{uv}\geq0$ (and if $s$ can be negative, then $w_{uv}$ should be in $[-1,1]$). One main advantage
of similarities is maybe to avoid (or minimize) problems with missing values setting them to 0.
\end{aside}

@@

Here we show why \autoref{p:edge_full} can model signed graphs as a special case, namely with $k=2$.
Remember the learning
bias we introduced in the previous chapter \vpageref{text:cc_new_bias}: every node belongs to
exactly one of the $K$ clusters available, is connected positively with the other nodes belonging to
the same cluster, and is connected negatively with the nodes belonging to every other clusters. Now
let the dimension of profiles be $d=K$ and the number of directions $k=2$. Moreover, we define the
two directions vectors as $w_1 = \frac{\onev}{\sqrt{d}} = w^+$ and $w_2 = -w_+ = w_-$, where
$\onev$ is the vector of dimension $d$ with only $1$ coordinates. Finally, for a node $u$ in
cluster $i$, we assume its profile $x_u$ is set as follow: all component are equal to $-b$ except
for the \ith{} one, which is equal to $a$, where $a$ and $b$ are constants we define now.

First, because $x_u$ is a unit vector, we have that $a^2 + (K-1)b^2=1$, from which $a =
\sqrt{1-(K-1)b^2}$ whenever $1-(K-1)b^2 \geq 0$, or equivalently $b \leq \frac{1}{\sqrt{K-1}}$.
Next, we look at $\onev^T\left(x_u \circ x_v\right) = \onev^T m_{uv} = \frac{1}{\sqrt{d}}
\sum_{i=1}^d m_{{uv}_i}$ for any edge $(u,v) \in E$. If $u$ and $v$ are in the same cluster, then
$\sum_{i=1}^d m_{{uv}_i} = a^2 + (K-1)b^2 = 1$, implying that $w_+^T m_{uv} = \frac{1}{d} > w_-^T
m_{uv} = -\frac{1}{d}$. Otherwise, $\sum_{i=1}^d m_{{uv}_i} = (K-2)b^2 - 2ab = (K-2)b^2 -
2b\sqrt{1-(K-1)b^2} = f_K(b)$. By computing the derivative of $f_K$ with respect to $b$, we find
that it reaches a minimum of $-\frac{1}{K-1}$ at $b = \frac{1}{\sqrt{K(K-1)}} \leq
\frac{1}{\sqrt{K-1}}$. In this case, $w_+^T m_{uv} = -\frac{1}{(K-1) \sqrt{d}} < w_-^T m_{uv} =
\frac{1}{(K-1) \sqrt{d}}$. To maximize $\sum_{(u,v)\in E} \max_{1 \leq \ell \leq d} w_\ell^T
m_{uv}$, we therefore assign $w_+$ to edges whose both endpoints belong to the same cluster, and
$w_-$ to edges joining two different clusters.

Moreover, given this assignment, we can show that $w_+$ and $w_-$ are the optimal directions, under
the additional uniformity assumption that there are exactly $r$ edges within every cluster, and $s$
edges between any two clusters. First, we define
\begin{equation*}
  m_{\mathrm{inner}} = \sum_{(u,v) \in E;\, \cluster(u)=\cluster(v)} m_{uv} \qquad \text{and} \qquad
  m_{\mathrm{outer}} = \sum_{(u,v) \in E;\, \cluster(u)\neq \cluster(v)} m_{uv}
\end{equation*}
to be the sum of all edge vectors within and between clusters respectively. By our choice of $a$ and
$b$, we have that $m_{\mathrm{inner}} = r\onev$ and $m_{\mathrm{outer}} = -\frac{s}{2}\onev$. Now
recall that the solution of $\argmax_{w\in \dsphere} w^T c$ is $\frac{c}{||c||}$. Therefore, the
best direction to assign to inner edges is indeed $w_+$, and so is $w_-$ to outer edges.

\begin{aside}
For inner edge vectors, $m_{uv}$ in cluster $i$ has $a^2$ as its \ith{} component, and $b^2$
elsewhere. There are $r$ of them in cluster $i$, so summing over all clusters we get
$m_{\mathrm{inner}} = r\left(a^2 + (K-1)b^2\right)\onev = r\onev$.
For outer edge vectors, there are a few more steps. Consider the $s$ edges between $C_1$ and $C_2$,
whose $m_{uv}$ is
$\begin{pmatrix}
  -ab \\
  -ab \\
  b^2 \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. Likewise, between $C_1$ and $C_3$ we have
$m_{uv}=\begin{pmatrix}
  -ab \\
  b^2 \\
  -ab \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. We thus we see that the first component is always $-ab$, that there is another
such term at the other cluster index, while the rest is $b^2$. Summing all the vectors incident to
$C_1$, we get
$s\begin{pmatrix}
  -ab(K-1) \\
  b^2(K-2) -ab \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$. Similarly, the sum of all vectors incident to $C_2$ is
$s\begin{pmatrix}
  b^2(K-2) -ab \\
  -ab(K-1) \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$, and so on for every cluster $i$. Summing over all clusters and dividing by $2$ to
avoid double counting, we get $m_{\mathrm{outer}} = \frac{s}{2}\left( (K-1)\left(b^2(K-2) -ab\right)
-ab(K-1) \right)\onev = \frac{s(K-1)}{2}\left(b^2(K-2) -2ab\right)\onev =
\frac{s(K-1)}{2}f_K(b)\onev = -\frac{s(K-1)}{2} \frac{1}{K-1} \onev = -\frac{s}{2}\onev$.
\end{aside}

@@

At the other extreme of $k=2$, when $k=|E|$, the topology of the graph becomes irrelevant,
as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} f(k)$ to increases as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers\stodo{elaborate on that connection}.

\stodo{move that in the conclusion}
\begin{aside}
Generalizing the balance theory from signed graphs, we could imagine that in triangles or short
cycles, only certain combinations of directions are valid (or at least desirable). As a nice side
effect, this would likely require more innovative optimization algorithms. However, it sounds very
application-dependant. Furthermore, in our current setting, we do not know in advance in the
semantic of each direction, and we have no labeled edges. Therefore, it is not clear how those
constraints would be specified or whether they can be learned, if applicable. 
\end{aside}
\stodo{Specification of constraints sounds related to network schema in \autocite[Definition
3]{HINSurvey17}, except we have a single node type and are looking at higher order patter.}

Another way to leverage the graph topology is to define local constraints at the node level. Since
the beginning we considered that the user profiles were given, but from now on we further assume
that they are normalized so that they are unit norm, just like directions. In practice, if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. One way to integrate that fact in
our optimization formulation is to assume that profiles are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real coefficients and $b_u
\in \Rbb^d$. Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it consume a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term $\sum_{u\in V} \left|\left| x_u - b_u -
\sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2$ to \eqref{eq:edge_full}.

@@

% $\theset{(u,v) \in E}{v \in \nei(u)}$

In order to make the optimization more tractable, we relax the formulation of \autoref{p:edge_full}
by replacing the $\max$ inside the sum of \eqref{eq:edge_full}, either by a continuous approximation
or by a low rank matrix approach.

One can check that given a set of numbers $S=\{a_1, a_2, \ldots, a_{|S|}\}$ and real number $\beta >
0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{\beta a_i} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{1}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}
Note that in second term, because we do not know the edge assignment, we have to decide which of the
$w_1, \ldots, w_k$ is $w_{uv}$ equal to. According to our definition of the goodness function, we
want to let $w_{uv} = \argmax_{w_\ell \in \mathcal{D}_k} {s_{uv}}^T w_\ell$. However, the $\argmax$
function is not convex and we thus use the same relaxation as before. Namely, by stacking all the
vectors of $\mathcal{D}_k$ into a single vector $W$ of size $k\times d$, the relaxation is:
$$ \begin{array}[t]{lrcl}
  \mathcal{A} : & \Rbb^m & \longrightarrow & \Rbb^d \\
      & (w_1,\ldots,w_k) & \longmapsto &
  \sum_{i=1}^k \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_i)} w_i
\end{array} $$
This function has an explicit derivative, but it is rather costly to compute and in practice we
found automatic differentiation~\autocite{autograd15} to be more efficient. While this modified
objective function is now jointly convex in all its variables, the set $\dsphere$ is not. To ease
the optimization procedure, we thus only impose the directions $w_\ell$ to lie within the unit ball
and expect them to have norm close to one in order to maximize the dot product in the ${s_{uv}}^T
w_\ell$ term.

Another way to remove the $\max$ is to first seek one direction for every edge, that is to maximize
$\sum_{u,v\in E} {s_{uv}}^T w_{uv}$. This objective can be written in matric form as $\tr(SW) =
\frop{S^T}{W}$, where $S \in \Rbb^{|E| \times d}$ is the $s_{uv}$ vectors for all edges stacked
vertically and $W \in \Rbb^{d \times |E|}$ is the matrix of all directions stacked horizontally.
More specifically, because of the unit norm constraint on directions, we look for $W$ in the set
$\mathbb{M}^{d\times |E|}$ of $d\times |E|$ matrices with unit $\ell_2$ norm columns. To avoid the
trivial solution of letting each direction $w_{uv}$ merely be $\frac{s_{uv}}{\|s_{uv}\|}$, we
further constraint $W$ to be of low rank. That is, what we really want to optimize is \[\min_{W\in
\mathbb{M}^{d\times |E|}} -\frop{S^T}{W} + \rank(W)\,.\] However, while the Frobenius inner product is
linear in $W$, the rank operator is highly irregular and optimizing it directly is \NPc{}. Therefore
we relax the rank term by the nuclear norm of $W$, defined as $\|W\|_* = \sum_{i=1}^{\min(d,|E|)}
\sigma_i(W)$, where $\sigma_i(W)$ is the \ith{} largest singular value of $W$. The intuition is
that the rank of $W$ is
equal to the number of non-zero singular value while the nuclear norm is essentially an $\ell_1$ norm on
the spectrum of $W$. Therefore, as minimizing the $\ell_1$ enforce sparsity, minimizing the nuclear
norm tends to make the rank smaller by ensuring the presence of zeros in the singular values of $W$.
\footnote{As mentioned in \autoref{sec:troll_related} about \autocite{OnlineCompletion17}, a tighter
relaxation of the low-rank constraint is the max norm, but is it convex?} This
results in the following optimization problem: \[\min_{W\in \mathbb{M}^{d\times |E|}} -\frop{S^T}{W}
+ \lambda \|W\|_*\,,\] where $\lambda$ is a regularization parameter. Since $\|\cdot\|_*$ is a norm, this
objective function is convex (although the domain over which we optimize is not), but it does not
have a gradient, forcing us to rely on potentially costly proximal gradient descent
methods~\autocite{Parikh2013a}. Instead, we consider the equivalent problem obtained by replacing
the regularization parameter with an upper bound $\delta$ on the nuclear norm of $W$, yielding:
\begin{equation*}
  \min_{\substack{W\in \mathbb{M}^{d\times |E|} \\ \|W\|_* \leq \delta}} -\frop{S^T}{W}\,.
\end{equation*}

>>>>

\section{Learning problem}
\label{sec:edge_problem}

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm\stodo{or bounded?}. As we will discuss, this
constraint can be seen as a way to couple the different dimensions of the profiles%
\stodo{don't remember what I meant here}.

@@

We assume in the following that both the profiles and the directions are vectors with components in
$[-1, 1]$. A natural choice for the operator $s$ is then the difference between the profiles, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_{uv}$, we expect
${x_u}^T w_{uv}$ to be close to ${x_v}^T w_{uv}$, as measured by $|{x_u}^T w_{uv} - {x_v}^T
w_{uv}|$. Because we prefer to work with smooth functions, we define the goodness to be $g(s_{uv},
w_{uv}) = \left({s_{uv}}^T w_{uv} \right)^2$, and we want to minimize it with respect to $w_{uv}$.
When on the \ith{} dimension ${x_u}_{;i} - {x_v}_{;i}$ is close to zero, the minimization objective
and the fact that $w_{uv}$ is
unit-norm impose that ${w_{uv}}_{;i}$ should be large. On the other hand, with this formulation,
there is no way to set ${w_{uv}}_{;i}$ in order to \enquote{highlight} the fact that
${x_u}_{;i}$ and ${x_v}_{;i}$ are widely different.
Whereas there might be situations where this is not a problem, as we discussed in the introduction,
here we seek to model heterophilic interactions, at least along a subset of dimensions. Therefore, in
the following, we will adopt a refined approach.

@@

\iffalse
\begin{aside}
When $s$ is a distance, and therefore positive, we impose $w_{uv}\geq 0$. A link $(u,v)$ is then
well explained by $w_{uv}$ when observed differences along this direction are small. More precisely,
when the quantity ${s_{uv}}^T w_{uv}$ is close to zero. Because $w_{uv}$ is unit norm, the \ith{}
component of $w_{uv}$ should be large when the \ith{} feature of $x_u$ and $x_v$ are close. Hence, a
null value for a component $i$ of $s_{uv}$ should not mean anything else than $u$ and $v$ share a
known and similar feature $i$ (\eg{} the unknown value should be different from $0$). When $s$ is
the difference $x_u - x_v$, it can be negative and the same reasoning can be applied, but with
$w_{uv}\in [-1,1]^d$ and a link value of $\left({s_{uv}}^T w_{uv}\right)^2$.

An alternative is to consider that the combination is a similarity. Links are created when the
observed similarity is high. Again, if $s$ is positive, it seems reasonable to impose that
$w_{uv}\geq0$ (and if $s$ can be negative, then $w_{uv}$ should be in $[-1,1]$). One main advantage
of similarities is maybe to avoid (or minimize) problems with missing values setting them to 0.
\end{aside}
\fi

@@

Here we show why \autoref{p:edge_full} can model signed graphs as a special case, namely with $k=2$
and specifically crafted profiles.
Remember the learning
bias we introduced in the previous chapter \vpageref{text:cc_new_bias}: every node belongs to
exactly one of the $K$ clusters available, is connected positively with the other nodes belonging to
the same cluster, and is connected negatively with the nodes belonging to every other clusters. Now
let the dimension of profiles be $d=K$ and the number of directions $k=2$. Moreover, we define the
two direction vectors as $w_1 = \frac{\onev}{\sqrt{d}} = w^+$ and $w_2 = -w_+ = w_-$, where
$\onev$ is the vector of dimension $d$ with only $1$ coordinates. Finally, for a node $u$ in
cluster $i$, we assume its profile $x_u$ is set as follow: all component are equal to $-b$ except
for the \ith{} one, which is equal to $a$, where $a$ and $b$ are real constants we define now.

First, because $x_u$ is a unit vector, we have that $a^2 + (K-1)b^2=1$, from which $a =
\sqrt{1-(K-1)b^2}$ whenever $1-(K-1)b^2 \geq 0$, or equivalently $b \leq \frac{1}{\sqrt{K-1}}$.
Next, we look at $\left(x_u \circ x_v\right)^T \onev = {s_{uv}}^T \onev = \frac{1}{\sqrt{d}}
\sum_{i=1}^d  {s_{uv}}_{;i}$ for any edge $(u,v) \in E$. If $u$ and $v$ are in the same cluster, then
$\sum_{i=1}^d  {s_{uv}}_{;i} = a^2 + (K-1)b^2 = 1$, implying that ${s_{uv}}^T w_+ = \frac{1}{d} >
{s_{uv}}^T w_- = -\frac{1}{d}$. Otherwise, $\sum_{i=1}^d  {s_{uv}}_{;i} = (K-2)b^2 - 2ab = (K-2)b^2 -
2b\sqrt{1-(K-1)b^2} = f_K(b)$. By computing the derivative of $f_K$ with respect to $b$, we find
that it reaches a minimum of $-\frac{1}{K-1}$ at $b = \frac{1}{\sqrt{K(K-1)}} \leq
\frac{1}{\sqrt{K-1}}$. In this case, ${s_{uv}}^T w_+ = -\frac{1}{(K-1) \sqrt{d}} < {s_{uv}}^T w_- =
\frac{1}{(K-1) \sqrt{d}}$. To maximize $\sum_{(u,v)\in E} \max_{1 \leq \ell \leq d} w_\ell^T
s_{uv}$, we therefore assign $w_+$ to edges whose both endpoints belong to the same cluster, and
$w_-$ to edges joining two different clusters.

Moreover, given this assignment, we can show that $w_+$ and $w_-$ are the optimal directions, under
the additional uniformity assumption that there are exactly $p$ edges within every cluster, and $q$
edges between any two clusters. First, we define
\begin{equation*}
  s_{\mathrm{inner}} = \sum_{(u,v) \in E;\, \cluster(u)=\cluster(v)} s_{uv} \qquad \text{and} \qquad
  s_{\mathrm{outer}} = \sum_{(u,v) \in E;\, \cluster(u)\neq \cluster(v)} s_{uv}
\end{equation*}
to be the sum of all edge vectors within and between clusters respectively. By our choice of $a$ and
$b$, we have that $s_{\mathrm{inner}} = p\onev$ and $s_{\mathrm{outer}} = -\frac{q}{2}\onev$. Now
recall that the solution of $\argmax_{w\in \dsphere} w^T c$ is $\frac{c}{||c||}$. Therefore, the
best direction to assign to inner edges is indeed $w_+$, and so is $w_-$ to outer edges.

\begin{aside}
For inner edge vectors, $s_{uv}$ in cluster $i$ has $a^2$ as its \ith{} component, and $b^2$
elsewhere. There are $p$ of them in cluster $i$, so summing over all clusters we get
$s_{\mathrm{inner}} = p\left(a^2 + (K-1)b^2\right)\onev = p\onev$.
For outer edge vectors, we first consider the $q$ edges between $C_1$ and $C_2$,
whose $s_{uv}$ is
$\begin{pmatrix}
  -ab \\
  -ab \\
  b^2 \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. Likewise, between $C_1$ and $C_3$ we have
$s_{uv}=\begin{pmatrix}
  -ab \\
  b^2 \\
  -ab \\
  b^2 \\
  \vdots \\  
\end{pmatrix}$. We thus we see that the first component is always $-ab$, that there is another
such term at the other cluster index, while the rest is $b^2$. Summing all the vectors incident to
$C_1$, we get
$q\begin{pmatrix}
  -ab(K-1) \\
  b^2(K-2) -ab \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$. Similarly, the sum of all vectors incident to $C_2$ is
$q\begin{pmatrix}
  b^2(K-2) -ab \\
  -ab(K-1) \\
  b^2(K-2) -ab \\
  \vdots \\  
\end{pmatrix}$, and so on for every cluster $i$. Summing over all clusters and dividing by $2$ to
avoid double counting, we get $s_{\mathrm{outer}} = \frac{q}{2}\left( (K-1)\left(b^2(K-2) -ab\right)
-ab(K-1) \right)\onev = \frac{q(K-1)}{2}\left(b^2(K-2) -2ab\right)\onev =
\frac{q(K-1)}{2}f_K(b)\onev = -\frac{q(K-1)}{2} \frac{1}{K-1} \onev = -\frac{q}{2}\onev$.
\end{aside}

@@

At the other extreme of having only $k=2$ directions, when $k=|E|$, the topology of the graph
becomes irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} f(k)$ to increase as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers. To retain a fair level of interpretability, we deem appropriate to have $k$ not
too large.

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level. 
Recall that we consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. One way to integrate that fact in
our optimization formulation is to assume that profiles are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real coefficients and $b_u
\in \Rbb^d$. Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it consume a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term 
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} = 
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}
from \eqref{eq:edge_full}.

@@

\section{Proposed approaches}
\label{sec:edge_methods}

After having introduced \autoref{p:edge_full} and some additional constraints, we now present
methods to solve it. We start with a naive baseline, and propose a post-processing to improve it.
Then we describe a convex relaxation. Finally, we depart slightly from finding $k$ directions by
considering a low-rank matrix formulation, where each edge is assigned a linear combination of a
small number of base directions, paving the way to overlapping clustering.

\subsection{\kmeans{} baseline and improvement}
\label{sub:edge_baseline}

\subsection{Vector optimization}
\label{sub:edge_vector}

The difficulty in optimizing directly \eqref{eq:edge_full} stem from the $\max$ operator, which we
replace here by a convex relaxation. Indeed, one can check that given a set of numbers $S=\{a_1,
a_2, \ldots, a_{|S|}\}$ and real number $\beta > 0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{\beta a_i} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{1}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}
Note that in second term, because we do not know the edge assignment, we have to decide which of the
$w_1, \ldots, w_k$ is $w_{uv}$ equal to. According to our definition of the goodness function, we
want to let $w_{uv} = \argmax_{w_\ell \in \mathcal{D}_k} {s_{uv}}^T w_\ell$. However, the $\argmax$
function is not convex and we thus use the same relaxation as before. Namely, by stacking all the
vectors of $\mathcal{D}_k$ into a single vector $W$ of size $k\times d$, the relaxation is:
$$ \begin{array}[t]{lrcl}
  \mathcal{A} : & \Rbb^m & \longrightarrow & \Rbb^d \\
      & (w_1,\ldots,w_k) & \longmapsto &
  \sum_{i=1}^k \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_i)} w_i
\end{array} $$
This function has an explicit derivative, but it is rather costly to compute and in practice we
found automatic differentiation~\autocite{autograd15} to be more efficient. While this modified
objective function is now jointly convex in all its variables, the set $\dsphere$ is not. To ease
the optimization procedure, we thus only impose the directions $w_\ell$ to lie within the unit ball
and expect them to have norm close to one in order to maximize the dot product in the ${s_{uv}}^T
w_\ell$ term.

\subsection{Matrix optimization}
\label{sub:edge_matrix}

Another way to remove the $\max$ is to first seek one direction for every edge, that is to maximize
$\sum_{u,v\in E} {s_{uv}}^T w_{uv}$. This objective can be written in matric form as $\tr(SW) =
\frop{S^T}{W}$, where $S \in \Rbb^{|E| \times d}$ is the $s_{uv}$ vectors for all edges stacked
vertically and $W \in \Rbb^{d \times |E|}$ is the matrix of all directions stacked horizontally
and $\frop{A}{B} = \tr\left(A^T B\right)$ is the Frobenius inner product of two real matrices.
More specifically, because of the unit norm constraint on directions, we look for $W$ in the set
$\mathbb{M}^{d\times |E|}$ of $d\times |E|$ matrices with unit $\ell_2$ norm columns. To avoid the
trivial solution of letting each direction $w_{uv}$ merely be $\frac{s_{uv}}{\|s_{uv}\|}$, we
further constraint $W$ to be of low rank. That is, what we really want to optimize is \[\min_{W\in
\mathbb{M}^{d\times |E|}} -\frop{S^T}{W} + \rank(W)\,.\] However, while the Frobenius inner product is
linear in $W$, the rank operator is highly irregular and optimizing it directly is \NPc{}. Therefore
we relax the rank term by the nuclear norm of $W$, defined as $\|W\|_* = \sum_{i=1}^{\min(d,|E|)}
\sigma_i(W)$, where $\sigma_i(W)$ is the \ith{} largest singular value of $W$. The intuition is
that the rank of $W$ is
equal to the number of non-zero singular value while the nuclear norm is essentially an $\ell_1$ norm on
the spectrum of $W$. Therefore, as minimizing the $\ell_1$ enforce sparsity, minimizing the nuclear
norm tends to make the rank smaller by ensuring the presence of zeros in the singular values of $W$.
\footnote{As mentioned in \autoref{sec:troll_related} about \autocite{OnlineCompletion17}, a tighter
convex relaxation of the low-rank constraint is the max norm, but we do not pursue that venue
further.} This
results in the following optimization problem: \[\min_{W\in \mathbb{M}^{d\times |E|}} -\frop{S^T}{W}
+ \lambda \|W\|_*\,,\] where $\lambda$ is a regularization parameter. Since $\|\cdot\|_*$ is a norm, this
objective function is convex (although the domain over which we optimize is not), but it does not
have a gradient, forcing us to rely on potentially costly proximal gradient descent
methods~\autocite{Parikh2013a}. Instead, we consider the equivalent problem obtained by replacing
the regularization parameter with an upper bound $\delta$ on the nuclear norm of $W$, yielding:
\begin{equation*}
  \min_{\substack{W\in \mathbb{M}^{d\times |E|} \\ \|W\|_* \leq \delta}} -\frop{S^T}{W}\,.
\end{equation*}
